Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2016-10-02 article.tex
theory max liklihood

\subsection{Maximum-likelihood method and non-linear optimization}

\paragraph{Parameter estimation} is a common task in statistics. It describes the process of inferring parameter values or parameter ranges of a statistical model based on observed data. Over decades, appropriate estimators have been developed for different problem classes. The principle of maximum likelihood allows to derive an estimator which is especially suited for applications where the distribution of the observation noise is known. This knowledge about the structure of the noise constitutes an additional information that makes the maximum-likelihood estimator (MLE) \textit{efficient}, i.e.~from all unbiased estimators the MLE has the lowest variance. Other properties of maximum-likelihood estimation are \textit{consistency}, i.e.~the true parameter is reached in the limit of infinite data-sample size, and \textit{asymptotic normality}, i.e.~in the limit of infinite data the MLE follows a multivariate normal distribution. See \citep{azzalini1996statistical} for an introduction to likelihood theory.

@@

The easiest and one of the most frequent situations is when $\phi\sim\mathcal N(\mu(\theta), \Sigma)$ follows a normal distribution. When measurements are statistically independent, the variance-covariance matrix $\Sigma$ is diagonal and $\phi$ factorizes. The Likelihood function becomes
\begin{align}
	L(\theta) &= \prod_i \frac{1}{\sqrt{2\pi\sigma_i^2}} e^{-\frac{\mu_i(\theta)^2}{2\sigma_i^2}} 
	\label{eq:likelihood}
\end{align}
The connection between the statistical model and the observed data is given by $\mu_i(\theta) = y_i(\theta) - y_i^D$. Taking twice the negative logarithm of eq.~\eqref{eq:likelihood} yields
\begin{align}
	l(\theta) = -2\log L(\theta) &= \sum_i \left[\left(\frac{y_i(\theta) - y_i^D}{\sigma_i}\right)^2 + \log(2\pi\sigma_i^2)\right].
	\label{eq:nll}
\end{align}
Assuming that the data uncertainties $\sigma_i$ are known, eq.~\eqref{eq:nll} is the weighted least-squares function shifted by a constant. For unknown $\sigma_i$, eq.~\eqref{eq:nll} can be optimized with respect to $\theta$ and $\sigma$ jointly, yielding the MLE $(\hat\theta, \hat\sigma)$.

\paragraph{Numerical optimization} is a diverse field with as many algorithms as there are optimization problems around.

>>>>

\subsection{Maximum-likelihood method}

Parameter estimation is a common task in statistics. It describes the process of inferring parameter values or parameter ranges of a statistical model based on observed data. Over decades, appropriate estimators have been developed for different problem classes. The principle of maximum likelihood allows to derive an estimator which is especially suited for applications where the distribution of the observation noise is known. This knowledge about the structure of the noise constitutes additional information that makes the maximum-likelihood estimator (MLE) \textit{efficient}, i.e.~from all unbiased estimators the MLE has the lowest variance. Other properties of maximum-likelihood estimation are \textit{consistency}, i.e.~the true parameter is reached in the limit of infinite data-sample size, and \textit{asymptotic normality}, i.e.~in the limit of infinite data the MLE follows a multivariate normal distribution. See \citep{azzalini1996statistical} for an introduction to likelihood theory.

@@

The easiest and one of the most frequent situations is when $\phi$ follows a normal distribution, $\phi\sim\mathcal N(\mu(\theta), \Sigma)$. When measurements are statistically independent, the variance-covariance matrix $\Sigma$ is diagonal and $\phi$ factorizes. The Likelihood function becomes
\begin{align}
	L(\theta) &= \prod_i \frac{1}{\sqrt{2\pi\sigma_i^2}} e^{-\frac{\mu_i(\theta)^2}{2\sigma_i^2}} 
	\label{eq:likelihood}
\end{align}
The connection between the statistical model and the observed data is given by $\mu_i(\theta) = y_i(\theta) - y_i^D$. Taking twice the negative logarithm of eq.~\eqref{eq:likelihood} yields
\begin{align}
	l(\theta) = -2\log L(\theta) &= \sum_i \left[\left(\frac{y_i(\theta) - y_i^D}{\sigma_i}\right)^2 + \log(2\pi\sigma_i^2)\right],
	\label{eq:nll}
\end{align}
thus converting the maximization of $L(\theta)$ to a minimization of $l(\theta)$. 
Assuming that the data uncertainties $\sigma_i$ are known, eq.~\eqref{eq:nll} is the weighted least-squares function shifted by a constant. For unknown $\sigma_i$, eq.~\eqref{eq:nll} can be optimized with respect to $\theta$ and $\sigma$ jointly, yielding the MLE $\hat\theta^{\prime} = (\hat\theta, \hat\sigma)$.

\subsection{Non-linear optimization}

Numerical optimization is a diverse field with as many algorithms as there are optimization problems around.
